quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-10-19 paper-quollr.Rmd
Added implementation workflow

```{r overview, eval=knitr::is_latex_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different \\twoD projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H'}
hex_grid_scurve + wrap_plots(
  scurve_proj_umap_model1, scurve_proj_umap_model2,
  scurve_proj_umap_model3, scurve_proj_umap_model4, 
  ncol = 2)
```

@@

The implementation of `quollr` is designed to be efficient, and easy to extend. The package is organized into a series of logical components that reflect the main stages of the workflow: data preprocessing, model fitting, low-density bin removal, prediction, visualization, and interactive exploration. This package structure makes the code easier to maintain and allows new features to be added without disrupting the existing functionality.

@@

Internally, `quollr` is organized into five primary modules that correspond to key steps in the analysis pipeline:

1. Data preparation — functions such as `gen_scaled_data()` standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
2. Hexagonal binning — the `hex_bin_data()` module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
3. Model fitting — the `fit_highd_model()` function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
4. Error computation — functions like `compute_errors()` and `aggregate_bin_error()` evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
5. Visualization and linkage — plotting and tour functions (`show_hex_model()`, `show_link_plots()`, and `show_error_link_plots()`) integrate outputs into linked static and dynamic displays using `ggplot2`, `langevitour`, and `crosstalk`.

@@

The internal data objects in `quollr` adhere to the **tidy data** principle, ensuring that every variable forms a column, every observation forms a row, and every type of observational unit forms a table.
This structure supports compatibility with the `tidyverse` and other R visualization frameworks.

The main object classes include:

* **`highd_data`** – a tibble containing the original high-dimensional observations with a unique identifier (`ID`) and variable columns prefixed with `"x"` (e.g., `x1`, `x2`, …).
* **`nldr_data`** – a tibble containing two-dimensional embeddings, labeled as `emb1` and `emb2`, matched to the same `ID`s.
* **`highd_vis_model`** – the primary model object returned by `fit_highd_model()`, containing two key components: the high-dimensional summaries for each bin and their corresponding two-dimensional positions.
* **`hex_bin_obj`** – the internal representation of the hexagonal grid, including bin coordinates (`a1`, `a2`), centroids, and normalized weights (`w_h`).

These objects provide a coherent foundation for data transfer across `quollr`’s modules—from binning to model fitting and visualization—supporting reproducibility and transparency in analyses.

@@

```{r, echo=TRUE}
bin_configs <- calc_bins_y(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 25, 
  q = 0.1)

@@

```{r, echo=TRUE}
hb_obj <- hex_binning(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 25, 
  q = 0.1)
```

@@

```{r, echo=TRUE}
all_centroids_df <- gen_centroids(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 25, 
  q = 0.1
  )

all_centroids_df
```

@@

```{r, echo=TRUE}
low_density_hex <- find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 25, 
  md_thresh = 0.05
)
```

@@

trimesh_limb <- ggplot() + 
  geom_point(
    data = tsne_limb_scaled_with_cluster,
    aes(
      x = emb1,
      y = emb2,
      color = cluster
    ),
    alpha = 0.3
  ) +
  geom_segment(data = trimesh_data_limb, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494')) +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )

trimesh_limb_int <- ggplotly(trimesh_limb, 
                                width = "350", 
                                height = "350", 
                                tooltip = "none") |>
  config(
    staticPlot = TRUE,        # Disables all interactivity (no hover, zoom, pan)
    displayModeBar = FALSE,   # Hides the plotly toolbar
    editable = FALSE,         # Disables annotations and editing
    showTips = FALSE,         # Removes tooltip on hover
    displaylogo = FALSE,      # Hides plotly logo
    responsive = FALSE        # Disables responsive resizing
  )

# hex_grid <- tsne_limb_obj$hb_obj$hex_poly
# counts_df <- tsne_limb_obj$hb_obj$std_cts
# 
# hex_grid_with_counts <- left_join(hex_grid, counts_df, by = c("h" = "h"))
# 
# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = n_h)) +
#   #geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = h)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed() +
#   theme_minimal()
```

@@

```{r langevitour-limb-tsne-author-proj}

data_limb_n <- data_limb_n |>
  select(-type) |>
  mutate(type = as.character(tsne_limb_scaled_with_cluster$cluster.ids))

df_model_data_limb_n <- bind_rows(df_b_limb, data_limb_n)

limb_highd_vis <- langevitour::langevitour(df_model_data_limb_n[1:(length(df_model_data_limb_n)-1)],
                         lineFrom = trimesh_data_limb$from_reindexed,
                         lineTo = trimesh_data_limb$to_reindexed,
                         group = factor(df_model_data_limb_n$type,
                                        c("0", "1", "2", "3", "4", "5", "6", "model")),
                         levelColors = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494', "#000000"),
                         enableControls = FALSE,
                         width = "421px", height = "421px")
```

@@

trimesh_limb_best_int <- ggplotly(trimesh_limb_best, 
                                width = "350", 
                                height = "350", 
                                tooltip = "none") |>
  config(
    staticPlot = TRUE,        # Disables all interactivity (no hover, zoom, pan)
    displayModeBar = FALSE,   # Hides the plotly toolbar
    editable = FALSE,         # Disables annotations and editing
    showTips = FALSE,         # Removes tooltip on hover
    displaylogo = FALSE,      # Hides plotly logo
    responsive = FALSE        # Disables responsive resizing
  )

# hex_grid <- tsne_best_limb_obj$hb_obj$hex_poly
# counts_df <- tsne_best_limb_obj$hb_obj$std_cts
# 
# hex_grid_with_counts <- left_join(hex_grid, counts_df, by = c("hex_poly_id" = "h"))
# 
# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = n_h)) +
#   #geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = h)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed() +
#   theme_minimal()
```

@@

```{r langevitour-limb-tsne-best-proj}

data_limb_n <- data_limb_n |>
  select(-type) |>
  mutate(type = as.character(tsne_limb_scaled_with_cluster$cluster.ids))

df_model_data_limb_n <- bind_rows(df_b_limb, data_limb_n)

limb_highd_vis_best <- langevitour::langevitour(df_model_data_limb_n[1:(length(df_model_data_limb_n)-1)],
                         lineFrom = trimesh_data_limb$from_reindexed,
                         lineTo = trimesh_data_limb$to_reindexed,
                         group = factor(df_model_data_limb_n$type,
                                        c("0", "1", "2", "3", "4", "5", "6", "model")),
                         levelColors = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494', "#000000"),
                         enableControls = FALSE,
                         width = "421px", height = "421px")
```

@@

```{r model-limb-html, eval=knitr::is_html_output(), fig.cap="Compare the published $2-\\text{D}$ layout (Figure \\@ref(fig:limb-rwbss) b) and the $2-\\text{D}$ layout selected (Figure \\@ref(fig:limb-rwbss) f) by RWBSS plot (Figure \\@ref(fig:limb-rwbss)) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$."}

modellimbfig <- htmltools::div(
        style = "display: grid; grid-template-columns: 1fr 1fr; row-gap: 0px; justify-items: center;",
        trimesh_limb_int,
        htmltools::div(style = "margin-top: 13px;", limb_highd_vis),
        trimesh_limb_best_int,
        htmltools::div(style = "margin-top: 13px;", limb_highd_vis_best)
      )

class(modellimbfig) <- c(class(modellimbfig), "htmlwidget")

modellimbfig
```

```{r}
df_exe_limb_int <- comb_all_data_model(
  highd_data = data_limb, 
  nldr_data = tsne_limb, 
  model_highd = df_bin_limb, 
  model_2d = df_bin_centroids_limb
)
```

@@

tsne_link_limb <- show_link_plots(
  point_data = df_exe_limb_int, 
  edge_data = trimesh_data_limb, 
  point_colour = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494')
)

class(tsne_link_limb) <- c(class(tsne_link_limb), "htmlwidget")

tsne_link_limb
```

```{r}
df_exe_limb_int_best <- comb_all_data_model(
  highd_data = data_limb, 
  nldr_data = tsne_limb2, 
  model_highd = df_bin_limb, 
  model_2d = df_bin_centroids_limb
)
```

@@

tsne_link_limb_best <- show_link_plots(
  point_data = df_exe_limb_int_best, 
  edge_data = trimesh_data_limb, 
  point_colour = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494')
)

class(tsne_link_limb_best) <- c(class(tsne_link_limb_best), "htmlwidget")

tsne_link_limb_best
```

>>>>

```{r overview, eval=knitr::is_latex_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different \\twoD projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H'}

hex_grid_scurve + wrap_plots(
  scurve_proj_umap_model1, scurve_proj_umap_model2,
  scurve_proj_umap_model3, scurve_proj_umap_model4, 
  ncol = 2)
```

@@

The implementation of `quollr` is designed to be efficient, and easy to extend. The package is organized into a series of logical components that reflect the main stages of the workflow: data preprocessing, model fitting, low-density bin removal, prediction, visualization, and interactive exploration (Figure `r '\\@ref(fig:workflow)'`). This package structure makes the code easier to maintain and allows new features to be added without changing the existing functionality.

```{r workflow, out.width = "100%", fig.pos="H", fig.cap="Workflow for "}

knitr::include_graphics("figures/quollr_workflow.png")
```

@@

The package is organized into seven core modules corresponding to stages of the analysis workflow: preprocessing, \twoD model construction, lifting into \pD, prediction, error computation, visualization, and interactivity. Each module performs a distinct task and communicates through data objects.

1. Data preprocessing — The function `gen_scaled_data()` standardizes the embedding data, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.

2. Construct \twoD model — `hex_binning()`, `merge_hexbin_centroids()`, `tri_bin_centroids()`, `gen_edges()`, `update_trimesh_index()`

3. Lift the model into \pD{} — the `avg_highd_data()`

4. Prediction - `predict_emb()`

5. Error computation — `glance()`

6. Visualization — `geom_hexgrid()`, `geom_trimesh()`, `show_langevitour()`

7. Interactivity - `show_link_plots()`, and `show_error_link_plots()`

@@

The internal data objects follow the tidy data principle: each variable is stored in a column, each observation in a row, and each type of information in its own table. This structure makes the package easy to use with the `tidyverse` and other R visualization tools.

### Input objects

- `highd_data` – a tibble containing the original high-dimensional observations with a unique identifier (`ID`) and variable columns prefixed with `"x"` (e.g., `x1`, `x2`, …).

- `nldr_data` – a tibble containing two-dimensional embeddings, labeled as `emb1` and `emb2`, matched to the same `ID`s.

### Generated objects

- `scaled_nldr_obj` — the output of `gen_scaled_data()`, which rescales the embedding to the range $[0, 1] \times [0, y_{2,\text{max}}]$, where $y_{2,\text{max}} = r_2 / r_1$ is the ratio of the embedding ranges. It includes the scaled coordinates (`scaled_nldr`) and the original limits (`lim1`, `lim2`).

- `hex_bin_obj` – the object created by `hex_binning()`, which defines the structure of the two-dimensional hexagonal grid used in modeling. It includes the grid spacing (`a1`, `a2`), the number of bins along each axis, the centroids of all hexagons, polygon coordinates for plotting, and the mapping of each data point to its assigned hexagon.

- `highd_vis_model` — the main model object returned by `fit_highd_model()`. It stores all components of the fitted visualization model, including the scaled NLDR data (`nldr_scaled_obj`), the hexagonal bin structure (`hb_obj`), the averaged \pD summaries for each bin (`model_highd`), the corresponding \twoD bin centroids (`model_2d`), and the triangulated mesh connecting neighboring bins (`trimesh_data`).

@@

```{r, echo=TRUE}
bin_configs <- calc_bins_y(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 21, 
  q = 0.1)

@@

```{r, echo=TRUE}
hb_obj <- hex_binning(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 21, 
  q = 0.1)
```

@@

```{r, echo=TRUE}
all_centroids_df <- gen_centroids(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 21, 
  q = 0.1
  )

head(all_centroids_df, 5)
```

@@

```{r, echo=TRUE}
low_density_hex <- find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 21, 
  md_thresh = 0.05
)
```

@@

trimesh_limb <- ggplot() + 
  geom_point(
    data = tsne_limb_scaled_with_cluster,
    aes(
      x = emb1,
      y = emb2,
      color = cluster
    ),
    alpha = 0.3
  ) +
  geom_segment(data = trimesh_data_limb, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494')) +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

@@

df <- inner_join(data_limb, tsne_limb_scaled_with_cluster, by = "ID")

### Define type column
df <- df |>
  select(starts_with("x"), starts_with("emb"), "cluster") |>
  mutate(type = tsne_limb_scaled_with_cluster$cluster) ## original dataset

df_b <- df_bin_limb |>
  filter(h %in% df_bin_centroids_limb$h) |>
  mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the h order in model_2d
df_b <- df_b[match(df_bin_centroids_limb$h, df_b$h),] |>
  tidyr::drop_na() |>
  select(-h)

df_exe_limb_int <- bind_rows(df_b, df)

point_data <- df_exe_limb_int 
edge_data <- trimesh_data_limb

num_highd_col <- point_data |>
  dplyr::select(starts_with("x")) |>
  NCOL()

shared_df <- crosstalk::SharedData$new(point_data)

nldr_plt <- shared_df |>
  ggplot(aes(x = emb1, y = emb2, colour = cluster)) +
  geom_point(alpha=0.5, size = 0.8) +
  geom_segment(data = trimesh_data_limb, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 0.5) +
  scale_color_manual(values = c("0" = '#66c2a5',"1" = '#fc8d62',"2" = '#8da0cb',"3" = '#e78ac3',"4" = '#a6d854',"5" = '#ffd92f',"6" = '#e5c494')) +
  theme_linedraw() +
  theme(
    #aspect.ratio = 1,
    plot.background = element_rect(fill = 'transparent', colour = NA),
    plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
    panel.background = element_rect(fill = 'transparent',
                                    colour = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank(),
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    legend.position = "none"
  )

langevitour_output <- langevitour::langevitour(point_data[1:num_highd_col],
                                               lineFrom = edge_data$from_reindexed,
                                               lineTo = edge_data$to_reindexed,
                                               group = factor(point_data$type,
                                                              c("0", "1", "2", "3", "4", "5", "6", "model")),
                                               pointSize = append(rep(point_sizes[1], NROW(df_b)),
                                                                  rep(point_sizes[2], NROW(df))),
                                               levelColors = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494', "#000000"),
                                               link=shared_df,
                                               linkFilter=FALSE)

linked_plt_int <- crosstalk::bscols(
  htmltools::div(style="display: grid; grid-template-columns: 1fr 1fr;",
                 nldr_plt,
                 htmltools::div(style = "margin-top: 20px;", langevitour_output)
  ),
  device = "xs"
)
```

@@

```{r}
df <- inner_join(data_limb, tsne_limb_scaled_with_cluster, by = "ID")

### Define type column
df <- df |>
  select(starts_with("x"), starts_with("emb"), "cluster") |>
  mutate(type = tsne_limb_scaled_with_cluster$cluster) ## original dataset

df_b <- df_bin_limb |>
  filter(h %in% df_bin_centroids_limb$h) |>
  mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the h order in model_2d
df_b <- df_b[match(df_bin_centroids_limb$h, df_b$h),] |>
  tidyr::drop_na() |>
  select(-h)

df_exe_limb_int <- bind_rows(df_b, df)

point_data <- df_exe_limb_int 
edge_data <- trimesh_data_limb

num_highd_col <- point_data |>
  dplyr::select(starts_with("x")) |>
  NCOL()

shared_df <- crosstalk::SharedData$new(point_data)

nldr_plt <- shared_df |>
  ggplot(aes(x = emb1, y = emb2, colour = cluster)) +
  geom_point(alpha=0.5, size = 0.8) +
  geom_segment(data = trimesh_data_limb, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 0.5) +
  scale_color_manual(values = c("0" = '#66c2a5',"1" = '#fc8d62',"2" = '#8da0cb',"3" = '#e78ac3',"4" = '#a6d854',"5" = '#ffd92f',"6" = '#e5c494')) +
  theme_linedraw() +
  theme(
    #aspect.ratio = 1,
    plot.background = element_rect(fill = 'transparent', colour = NA),
    plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
    panel.background = element_rect(fill = 'transparent',
                                    colour = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank(),
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    legend.position = "none"
  )

langevitour_output <- langevitour::langevitour(point_data[1:num_highd_col],
                                               lineFrom = edge_data$from_reindexed,
                                               lineTo = edge_data$to_reindexed,
                                               group = factor(point_data$type,
                                                              c("0", "1", "2", "3", "4", "5", "6", "model")),
                                               pointSize = append(rep(point_sizes[1], NROW(df_b)),
                                                                  rep(point_sizes[2], NROW(df))),
                                               levelColors = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494', "#000000"),
                                               link=shared_df,
                                               linkFilter=FALSE)

linked_plt_best <- crosstalk::bscols(
  htmltools::div(style="display: grid; grid-template-columns: 1fr 1fr;",
                 nldr_plt,
                 htmltools::div(style = "margin-top: 20px;", langevitour_output)
  ),
  device = "xs"
)
```

@@

@@

@@

class(linked_plt_int) <- c(class(linked_plt_int), "htmlwidget")

linked_plt_int
```

@@

class(linked_plt_best) <- c(class(linked_plt_best), "htmlwidget")

linked_plt_best
```
